\MakeUppercase{The Structure of $k$-Planar Graphs}

### 2019-07-05 k-planar.tex
Careful check of calculations

\begin{thm}
\label{k-planar}
Every $k$-planar graph has a layered $H$-partition of layered width at most $15k^2 + 30k$ in which $H$ has treewidth at most $\treewidth$.
\end{thm}

@@

\begin{clm}\clmlabel{generalized-tripod}
  For each $x\in V(T)$, let $V_x = \bigcup_{y\in V(T_x)} Y_y$, $F_x=\{w\in V(G): vw\in E(G), v\in V_x,\, w\not\in V_x\}$, and $N_x=V_x\cup F_x$.  Then, 
  \begin{compactenum}[(Y1)]
    \item\ylabel{thickness} $\mathcal{Y}:=(Y_x: x\in V(T))$ is a partition of $V(G)$ of layered width (with respect to $\mathcal{L}$) at most $\ell$.
    \item\ylabel{separator} For every $x\in V(T)$, there is no edge $vw\in E(G)$ with $v\in V_x$ and $w\in V(G)\setminus N_x$. 
    \item\ylabel{ancestor-edge} For every $x\in V(T)$, $x$ has $t'\le t$ strict $T$-ancestors $a_1,\ldots,a_{t'}$ such that $F_x \subseteq \bigcup_{i=1}^{t'} Y_{a_i}$.
  \end{compactenum}
\end{clm}

@@

  To establish Property~\yref{ancestor-edge}, consider some vertex $w\in F_x$.  Since $w\in F_x$, there exists an edge $vw\in E(G)$ with $v\in V_x$ and $w\not\in V_x$.  Since $v\in V_x$, $v\in Y_{x'}=Q_{x'}$ for some $T$-descendant $x'$ of $x$ (possibly $x=x'$). Since $\mathcal{Y}$ is a partition, $w\in Y_{a}=Q_a$ for some $a\not\in V(T_x)$.  Since $vw\in E(G)$, $x'a\in E(H)$.  By \tref{ancestor-edge}, one of $a$ or $x'$ is a $T$-ancestor of the other. Since $w\in Y_a\subseteq V_a$ and $w\not\in V_x\supseteq V_{x'}$, \yref{containment-i} rules out the possibility that $x'$ is a $T$-ancestor of $a$. Therefore, $a$ is a $T$-ancestor of $x$ which is a $T$-ancestor of $x'$.  Let $z_0,\ldots,z_r$ be the path, in $T$ from $z_0:=x'$ to $z_r:=a$.  For each $i\in\{0,\ldots,r\}$, at least one of $a$ or $x'$ is in $B_{z_i}$.  However, by \tref{subtree-root} $x'$ is not contained in $B_{x_i}$ for any $i\in\{1,\ldots,r\}$.  Therefore $a\in B_{x_i}$ for each $i\in\{0,\ldots,r\}$.  In particular, $a$ is contained in $B_x$.
  Property~\yref{ancestor-edge} now follows from the fact that $|B_x|\le t+1$ and $C_x$ contains $x$.
\end{proof}

@@

  Let $H^+$ denote the supergraph of $H$ with vertex set $V(T)$ and in which $xy\in E(H^+)$ if and only there exists some $z\in V(T)$ such that $x,y\in B_z$. 
  % Note that, by \tref{ancestor-edge}, $xy\in E(H^+)$ if and only if $x\in B_y$ or $y\in B_x$.
  We claim that $a_0,\ldots,a_p$ is a lazy walk\footnote{A \emph{lazy walk} in a graph $H$ is a walk in the non-simple graph $H'$ obtained by adding a self loop to each vertex of $H$.} in $H^+$.  Indeed, if $a_i\neq a_{i+1}$ for some $i\in\{0,\ldots,p-1\}$ then this is precisely because $w_i\in V_{a_i}$ but $w_{i+1}\not\in V_{a_i}$.  By definition, $w_i\in Y_{a_i'}$ for some $T$-descendant $a_i'$ of $a_i$.
  By \yref{separator}, $w_{i+1}\in F_{a_i}$ so by \yref{ancestor-edge} $w_{i+1}\in Y_{a_i''}$ for some strict $T$-ancestor $a_i''$ of $a_i$.  Since $w_iw_{i+1}\in E(G)$, $a_i'a_i''\in E(H)$.  By \tref{subtree-root}, $a_i''\in B_{a_i''}$ and $a_i''\in B_{a_i'}$.  Since $a_i$ is on the path from $a_i'$ to $a_i''$ in $T$ this implies that $a_i''\in B_{a_i}$.  Therefore $a_ia_i''\in E(H^+)$ as claimed.

@@

We can reduce the layered width of the $H$-partition of $G$ from $O(k^3)$ to $O(k^2)$ by observing that the dummy vertices in $V(G_0)\setminus V(G)$ do not contribute to the layered width of this partition.  In this setting, the proof of \cref{clm:general-width} is simpler since each vertex $w\in Y_x$ contributes at most two vertices to $V_i\cap Y_x$.  More precisely, a path $P\in\mathcal{P}$ containing an internal (dummy) vertex $w\in Y_x\cap (V_{i-j}\cup V_{i+j})$
\begin{enumerate}
  \item Contributes at most 2 vertices to $S_x\cap V_i$ for $j\in\{0,\ldots,\floor{(k+1)/2}$;
  \item Contributes at most 1 vertex to $S_x\cap V_j$ for $j\in\{\floor{(k+1)/2},\ldots,k+1\}$; 
  \item Contributes no vertices to $S_x\cap V_j$ for $j > k+1$.
\end{enumerate}
Redoing the calculation at the end of the proof of \cref{clm:general-width} then yields $|S_x\cap Y_i|\le d\ell(5k/2+5)=15k+30$.  With this change, the layered width of the partition given by \cref{modern-major-general} becomes $15k^2+30k$.  This establishes \cref{k-planar}.

@@

\begin{thm}
\label{g-k-planar}
Every $(g,k)$-planar graph has a layered $H$-partition of layered width at most 
$\max\{2g,3\}(5k^2 + 10k)$ in which $H$ has treewidth at most $\binom{k+5}{4}$.
\end{thm}

\note{PM}{I remove the $(g,1)$-specialization. It requires cutting surfaces.  For reference, the statement was:

@@

>>>>

\begin{thm}
\label{k-planar}
Every $k$-planar graph has a layered $H$-partition of layered width at most $18k^2 + 30k$ in which $H$ has treewidth at most $\treewidth$.
\end{thm}

@@

\begin{clm}\clmlabel{generalized-tripod}
  For each $x\in V(T)$, let $V_x := \bigcup_{y\in V(T_x)} Y_y$, $F_x:=\{w\in V(G): vw\in E(G), v\in V_x,\, w\not\in V_x\}$, and $N_x:=V_x\cup F_x$.  Then, 
  \begin{compactenum}[(Y1)]
    \item\ylabel{thickness} $\mathcal{Y}:=(Y_x: x\in V(T))$ is a partition of $V(G)$ of layered width (with respect to $\mathcal{L}$) at most $\ell$.
    \item\ylabel{separator} For each $x\in V(T)$, there is no edge $vw\in E(G)$ with $v\in V_x$ and $w\in V(G)\setminus N_x$. 
    \item\ylabel{ancestor-edge} For each $x\in V(T)$, there is a set $\{a_1,\ldots,a_{t'}\}$ of $t'\le t$ strict $T$-ancestors of $x$ such that $F_x \subseteq \bigcup_{i=1}^{t'} Y_{a_i}$.
  \end{compactenum}
\end{clm}

@@

  To establish Property~\yref{ancestor-edge}, consider some vertex $w\in F_x$.  Since $w\in F_x$, there exists an edge $vw\in E(G)$ with $v\in V_x$ and $w\not\in V_x$.  Since $v\in V_x$, $v\in Y_{x'}$ for some $T$-descendant $x'$ of $x$ (possibly $x=x'$). Since $\mathcal{Y}$ is a partition, $w\in Y_{a}$ for some $a\not\in V(T_x)$.  Since $vw\in E(G)$, we have $x'a\in E(H)$.  By \tref{ancestor-edge}, one of $a$ or $x'$ is a $T$-ancestor of the other. Since $w\in Y_a\subseteq V_a$ and $w\not\in V_x\supseteq V_{x'}$, \yref{containment-i} rules out the possibility that $x'$ is a $T$-ancestor of $a$. Therefore, $a$ is a $T$-ancestor of $x$ which is a $T$-ancestor of $x'$.  Let $z_0,\ldots,z_r$ be the path in $T$ from $z_0:=x'$ to $z_r:=a$.  For each $i\in\{0,\ldots,r\}$, at least one of $a$ or $x'$ is in $B_{z_i}$.  However, by \tref{subtree-root} $x'$ is not contained in $B_{x_i}$ for any $i\in\{1,\ldots,r\}$.  Therefore $a\in B_{x_i}$ for each $i\in\{0,\ldots,r\}$.  In particular, $a$ is contained in $B_x$.
  Property~\yref{ancestor-edge} now follows from the fact that $|B_x|\le t+1$ and $B_x$ contains $x$.
\end{proof}

@@

  Let $H^+$ denote the supergraph of $H$ with vertex set $V(T)$ and in which $xy\in E(H^+)$ if and only there exists some $z\in V(T)$ such that $x,y\in B_z$. 
  % Note that, by \tref{ancestor-edge}, $xy\in E(H^+)$ if and only if $x\in B_y$ or $y\in B_x$.
  We claim that $a_0,\ldots,a_p$ is a lazy walk\footnote{A \emph{lazy walk} in a graph $H$ is a walk in the pseudograph $H'$ obtained by adding a self loop to each vertex of $H$.} in $H^+$.  Indeed, if $a_i\neq a_{i+1}$ for some $i\in\{0,\ldots,p-1\}$ then this is precisely because $w_i\in V_{a_i}$ but $w_{i+1}\not\in V_{a_i}$.  By definition, $w_i\in Y_{a_i'}$ for some $T$-descendant $a_i'$ of $a_i$.
  By \yref{separator}, $w_{i+1}\in F_{a_i}$ so by \yref{ancestor-edge} $w_{i+1}\in Y_{a_i''}$ for some strict $T$-ancestor $a_i''$ of $a_i$.  Since $w_iw_{i+1}\in E(G)$, $a_i'a_i''\in E(H)$.  By \tref{subtree-root}, $a_i''\in B_{a_i''}$ and $a_i''\in B_{a_i'}$.  Since $a_i$ is on the path from $a_i'$ to $a_i''$ in $T$ this implies that $a_i''\in B_{a_i}$.  Therefore $a_ia_i''\in E(H^+)$ as claimed.

@@

We can reduce the layered width of the $H$-partition of $G$ from $O(k^3)$ to $O(k^2)$ by observing that the dummy vertices in $V(G_0)\setminus V(G)$ do not contribute to the layered width of this partition.  In this setting, the proof of \cref{clm:general-width} is simpler since each vertex $w\in Y_x$ contributes at most two vertices to $V_i\cap Y_x$.  More precisely, a path $P\in\mathcal{P}$ containing an internal (dummy) vertex $w\in Y_x\cap (V_{i-j}\cup V_{i+j})$
\begin{enumerate}
  \item Contributes at most 2 vertices to $S_x\cap V_i$ for $j\in\{0,\ldots,\floor{(k+1)/2}$;
  \item Contributes at most 1 vertex to $S_x\cap V_j$ for $j\in\{\floor{(k+1)/2},\ldots,k+1\}$; 
  \item Contributes no vertices to $S_x\cap V_j$ for $j > k+1$.
\end{enumerate}
Redoing the calculation at the end of the proof of \cref{clm:general-width} then yields 
\begin{align*}
    |S_x\cap Y_i| &\le d\ell\left( 
       2 
       + 4\left\lfloor\frac{k+1}{2}\right\rfloor 
       + 2\left(k+1-\left\lfloor\frac{k+1}{2}\right\rfloor\right)
       \right)\\
       & =
       d\ell\left(
         2k + 2\left\lfloor\frac{k+1}{2}\right\rfloor + 4
       \right) \\
       & \le
       d\ell(3k+5) = 18k+30 \enspace .
\end{align*}    
With this change, the layered width of the partition given by \cref{modern-major-general} becomes $18k^2+30k$.  This establishes \cref{k-planar}.

@@

\begin{thm}
\label{g-k-planar}
Every $(g,k)$-planar graph has a layered $H$-partition of layered width at most 
$\max\{2g,3\}\cdot(6k^2 + 10k)$ in which $H$ has treewidth at most $\binom{k+5}{4}$.
\end{thm}

\note{PM}{I removed (commented out) the $(g,1)$-specialization. It requires cutting surfaces.  For reference, the statement was:

@@

\note{PM}{By now this rough characterization is closely related to $(k+1,2)$-shortcut systems.  Let's think about how to present this.}
